---
title: "P3_PharmacovigilanceData_2"
author: "Dominik Frei"
date: "24 3 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rvest)
setwd("C:/Users/domin/Desktop/Dominik/DS/Projects/P3_PharmacovigilanceData")
```
P3 PharmacovigilanceData Part 2

As I was able to set up a data base in which I can import the XML files
provided by the FDA. I will now download more data (all files for the years
2021 and 2020), unzip them and write the insert statements for MySQL
in this R Script.


create the URLs to the files.
The format is: https://fis.fda.gov/content/Exports/faers_xml_YYYYQQ.zip
```{r}
years <- c("2021", "2020")
quarters <- c("Q1", "Q2", "Q3", "Q4")
URLs <- list()
ids <- list()

for (i in 1:2) {
        for (j in 1:4) {
                id <- paste(years[i], 
                            quarters[j],
                            sep = "")
                ids <- c(id, ids)
                
                URL <- paste("https://fis.fda.gov/content/Exports/faers_xml_",
                             id,
                             ".zip",
                             sep = "")
                URLs <- c(URL, URLs)
        }
}
```

download the files
```{r}
for (i in 1:length(URLs)) {
        dest <- paste("FAERS_XML/files/", 
                      ids[[i]], 
                      ".zip", 
                      sep = "")
        download.file(url = URLs[[i]],
                      destfile = dest)
}
```
unzip the files
```{r}
for (i in 1:length(URLs)) {
        file <- paste("FAERS_XML/files/", 
                      ids[[i]], 
                      ".zip", 
                      sep = "")
        directory <- "FAERS_XML/files/unziped"
        unzip(zipfile = file, exdir = directory, overwrite = TRUE)
}
```

write the insert statements for MySQL:

table: "general":
LOAD XML LOCAL INFILE 'C:/ProgramData/MySQL/MySQL Server 8.0/Uploads/X_ADRYYQQ.xml'
INTO TABLE `general`
ROWS IDENTIFIED BY '<safetyreport>';

table: "drugs":
LOAD XML LOCAL INFILE 'C:/ProgramData/MySQL/MySQL Server 8.0/Uploads/X_ADRYYQQ.xml'
INTO TABLE drugs
ROWS IDENTIFIED BY '<drug>';

table: "reactions":
LOAD XML LOCAL INFILE 'C:/ProgramData/MySQL/MySQL Server 8.0/Uploads/X_ADRYYQQ.xml'
INTO TABLE reactions
ROWS IDENTIFIED BY '<reaction>';

```{r}
#define necessary names for loop
table_names <- c("`general`", "drugs", "reactions")
identified_by <- c("'<safetyreport>'", "'<drug>'", "'<reaction>'")
file_numbers <- c("1", "2", "3")
years_short <- c("20", "21")
#quarters has already been defined

statements <- list()
outp <- list()

for (t in 1:3) {
        table <- table_names[[t]]
        identifyer <- identified_by[[t]]
        for (y in 1:2) {
                year <- years_short[[y]]
                for (f in 1:3) {
                        file <- file_numbers[[f]]
                        for (q in 1:4) {
                                quart <- quarters[[q]]
                                filename <- paste(file, "_ADR", year, quart, ".xml", sep = "")
                                statements <- c(statements, paste("LOAD XML LOCAL INFILE 'C:/ProgramData/MySQL/MySQL Server 8.0/Uploads/",
                                                                  filename, 
                                                                  "'\n",
                                                                  "INTO TABLE ",
                                                                  table,
                                                                  "\nROWS IDENTIFIED BY ",
                                                                  identifyer,
                                                                  ";\n\n",
                                                                  sep = ""))
                                                                  
                                
                        }
                                
                }
        }
        
}

for (i in 1:length(statements)) {
        cat(statements[[i]])
}
```

This output was copied into the SQL Script (P3_PharmacovigilanceData_2.sql) and
run sucessfully.